---
title: "Assignment 2: Coding Basics"
author: "Jack Carpenter"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
#create a sequence from 1:100 by 4s
seq(1,100,4)
#name the sequence
fours_sequence <- seq(1,100,4)
fours_sequence
#2. 
#run median and mean of fours_sequence
median(fours_sequence)
mean(fours_sequence)
#3. 
#compare the mean and median values - is mean or median greater?
mean(fours_sequence)>median(fours_sequence)
#the mean is NOT greater than the median (equal) so this comes back as FALSE
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#create a vector with student names
student_name<- c("Ted", "Beard", "Keeley", "Roy")
student_name
#create a vector with test scores
test_score <- c(98,87,69,48)
test_score
#create a vector whether they passed or not
passed <- c(test_score>50)
passed
#combine vectors into dataframe
student_scores <- data.frame(student_name, test_score, passed)
student_scores
#column names already appear to be labeled usefully
class(student_scores)
#double-check that its actually a dataframe
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: I don't know, what's a matrix? 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
